---
title: "Code for Liu etal"
author: "Mo Liu"
date: "15/05/2020"
output: html_document
---

It is difficult to provide everything for the whole pipeline, as some data is downloaded from QBiC website and some data is stored locally. The code is here for reference. 
Please feel free to email me if you are interested in any part of this study and I would like to provide more details about processing data.

required packages
```{r}
library(gplots)
library(RColorBrewer)
library(data.table)
library(enrichR)
library(ggplot2)
library(devtools)
library(stringi)
library('lsa') # for cosine()

```


```{r}
unartifact.signatures <- c("SBS1",  "SBS2" ,  "SBS3" ,  "SBS4"  , "SBS5" ,"SBS6",   "SBS7a" , "SBS7b",  "SBS7c",  "SBS7d" , "SBS8", "SBS9","SBS10a" ,"SBS10b","SBS11" , "SBS12" , "SBS13",  "SBS14",  "SBS15",  "SBS16", "SBS17a", "SBS17b", "SBS18",  "SBS19",  "SBS20", "SBS21" , "SBS22" , "SBS23" , "SBS24", "SBS25", "SBS26", "SBS28",  "SBS29",  "SBS30"  ,"SBS31" , "SBS32",  "SBS33" , "SBS34" , "SBS35",  "SBS36" , "SBS37" , "SBS38" , "SBS39"  ,"SBS40" , "SBS41", 
 "SBS42" , "SBS44" )

```


Overview of TF

I used Signature-QBiC to generate GRs and LRs for all PBMs with all unartifact signatures. Format is PBM - GR - LR.

```{r}
#TF.uPBM.info is read from TableS4
TF.median.GRs <- data.frame(TF.uPBM.info$Gene)
TF.median.LRs <- data.frame(TF.uPBM.info$Gene)


for(i in 1:nrow(TF.uPBM.info)){
  
  TF.uPBMs <- unlist(strsplit(TF.uPBM.info$PBMs[i],";"))
  
  for(signature in unartifact.signatures){
    
    
    output.name.genome <- paste("PCAWG.",signature,".genome.mean.ratio.matrix",sep="")
    
    matrix.genome <- get(output.name.genome)
    

    TF.median.GRs[i,signature] <- median(matrix.genome$LR[which(matrix.genome$uPBM %in% TF.uPBMs)])
    TF.median.LRs[i,signature] <- median( matrix.genome$GR[which(matrix.genome$uPBM %in% TF.uPBMs)])
    
    
  }
  
}



all.TF.LRs <- as.matrix(TF.median.LRs[,-1]) ##LRs, the first column is TF name
all.TF.GRs <- as.matrix(TF.median.GRs[,-1]) ##GRs

row.names(all.TF.LRs) <- TF.uPBM.info$Gene ##data1 and data2 can be found in 
row.names(all.TF.GRs) <- TF.uPBM.info$Gene

data <- t(as.matrix(rbind(-all.TF.LRs,all.TF.GRs)))


col = c("#081D58", "#253494","#225EA8","#1D91C0","#41B6C4","#A1DAB4","#FFFFCC","#FFFFCC","#FED976","#FEB24C","#FD8D3C","#FC4E2A","#E31A1C","#800026")
##heatmap in Figure 3
hm <- heatmap.2(data,trace="none",col = c("#081D58", "#253494","#225EA8","#1D91C0","#41B6C4","#FFFFCC","#FFFFCC","#FEB24C","#FD8D3C","#FC4E2A","#E31A1C","#800026"),
            hclustfun = function(x) hclust(x,method = "complete"),
            hline =TRUE,vline=TRUE,symkey=F,breaks=c(-3.5,-3,-2.5,-2,-1.5,-1,0,1,1.5,2,2.5,3,3.5),
            dendrogram = "both",
            labCol = NA,
            tracecol = "red",
            lhei=c(0.5,4), lwid=c(0.2,3.5), keysize=0.75, key.par = list(cex=0.5))  


TF.gain.loss.binding.number <- data.frame(unartifact.signatures) #count number of TFs with GR or LR >1

for(signature in unartifact.signatures){
  
  TF.gain.loss.binding.number$gain.TF.numbers[which(TF.gain.loss.binding.number[,1]==signature)] <- length(unique(TF.median.GRs$Gene[TF.median.GRs[,signature]>1]))
  
  TF.gain.loss.binding.number$loss.TF.numbers[which(TF.gain.loss.binding.number[,1]==signature)] <- length(unique(TF.median.LRs$Gene[TF.median.LRs[,signature]>1]))
  
  
}

plot.signature.TF.summary <- data.frame(c(TF.gain.loss.binding.number$gain.TF.numbers,
                                          TF.gain.loss.binding.number$loss.TF.numbers))

plot.signature.TF.summary$signature <- c(unartifact.signatures,unartifact.signatures)

plot.signature.TF.summary$category <- rep(c("TFs gaining binding affinity", "TFs abrogating binding affinity"),each=47)

colnames(plot.signature.TF.summary)[1] <- "TF.numbers"

plot.signature.TF.summary.gain <- plot.signature.TF.summary[1:47,]

plot.signature.TF.summary.gain <- plot.signature.TF.summary.gain[order(plot.signature.TF.summary.gain$TF.numbers,decreasing=T),]

##barplot in Figure 3

ggplot(data=plot.signature.TF.summary.gain, aes(x=signature, y=TF.numbers, fill=category)) +
  geom_bar(stat="identity")+
  scale_fill_manual("legend", values = c("TFs gaining binding affinity" = "#FC4E2A", "TFs abrogating binding affinity" = "#1D91C0"))+
  theme_bw()+theme(text = element_text(size=20),
                   axis.text.x = element_text(angle=90, hjust=1))   +scale_x_discrete(limits=plot.signature.TF.summary.gain$signature)+ theme(legend.position = "none")


plot.signature.TF.summary.loss <- plot.signature.TF.summary[48:94,]
plot.signature.TF.summary.loss <- plot.signature.TF.summary.loss[order(plot.signature.TF.summary.loss$TF.numbers,decreasing=T),]


ggplot(data=plot.signature.TF.summary.loss, aes(x=signature, y=TF.numbers, fill=category)) +
  geom_bar(stat="identity")+
  scale_fill_manual("legend", values = c("TFs gaining binding affinity" = "#FC4E2A", "TFs abrogating binding affinity" = "#1D91C0"))+
  theme_bw()+theme(text = element_text(size=20),
                   axis.text.x = element_text(angle=90, hjust=1))   +scale_x_discrete(limits=plot.signature.TF.summary.loss$signature)+ theme(legend.position = "none")
dev.off()
```

##TF.DBD in each cluster Figure 3

```{r}
hc.cutree.for.TF.clustering <- cutree(as.hclust(hm$colDendrogram), 1:1164)

cluster.A.TF <- names(hc.cutree.for.TF.clustering[hc.cutree.for.TF.clustering[,4]==2,4])
cluster.B.TF <- names(hc.cutree.for.TF.clustering[hc.cutree.for.TF.clustering[,4]==1,4])
cluster.C.TF <- names(hc.cutree.for.TF.clustering[hc.cutree.for.TF.clustering[,4]==4,4])
cluster.D.TF <- names(hc.cutree.for.TF.clustering[hc.cutree.for.TF.clustering[,4]==3,4])


cluster.A.TF <- data.frame(cluster.A.TF)

cluster.A.TF$cluster <- "A"

cluster.A.TF$TF.family <- TF.uPBM.info$DBD[match(cluster.A.TF[,1],TF.uPBM.info$Gene)]

cluster.B.TF <- data.frame(cluster.B.TF)

cluster.B.TF$cluster <- "B"

cluster.B.TF$TF.family <- TF.uPBM.info$DBD[match(cluster.B.TF[,1],TF.uPBM.info$Gene)]


cluster.C.TF <- data.frame(cluster.C.TF)

cluster.C.TF$cluster <- "C"

cluster.C.TF$TF.family <- TF.uPBM.info$DBD[match(cluster.C.TF[,1],TF.uPBM.info$Gene)]
cluster.D.TF <- data.frame(cluster.D.TF)

cluster.D.TF$cluster <- "D"

cluster.D.TF$TF.family <- TF.uPBM.info$DBD[match(cluster.D.TF[,1],TF.uPBM.info$Gene)]

##organize all clustering information together
cluster.composition.summary <- data.frame(matrix(nrow=0,ncol=4))

for(cluster.name in c("cluster.A.TF","cluster.B.TF","cluster.C.TF",
                      "cluster.D.TF")){
  
  cluster <- get(cluster.name)
  
  colnames(cluster)[1] <- "TFname"
  
  cluster <- cluster[!duplicated(cluster$TFname),]
  
  cluster.composition.matrix <- data.frame(table(cluster$TF.family))
  
  
  
  cluster.composition.matrix$cluster <- cluster.name
  
  cluster.composition.matrix$percentage <- cluster.composition.matrix$Freq/sum(cluster.composition.matrix$Freq)
  
  cluster.composition.summary <- rbind(cluster.composition.summary,cluster.composition.matrix)
  
  
}

```



```{r}
dbs <- listEnrichrDbs()
dbs.selected <- c("Reactome_2016")

for(signature in unartifact.signatures){
  
  gain.TF.list <- TF.median.GRs$Gene[TF.median.GRs[,signature]>1] ##selected TFs with GR>1
  
  PathwaySelection(gain.TF.list,0.005,"Reactome_2016")
  
  loss.TF.list <- TF.median.LRs$Gene[TF.median.LRs[,signature]>1] ##selected TFs with LR>1
  
  PathwaySelection(loss.TF.list,0.005,"Reactome_2016")
}




pathways.summary.gain.binding <- data.frame(matrix(nrow=0,ncol=13))
pathways.summary.loss.binding <- data.frame(matrix(nrow=0,ncol=13))


for(signature in unartifact.signatures){
  
  
  
  
  outname.enrichr.gain <- paste("filtered.median.enrichR.Reactome.gained.by.",signature,sep="")
  
  outname.enrichr.loss <- paste("filtered.median.enrichR.Reactome.lossed.by.",signature,sep="")
  
  gain.matrix <- get(outname.enrichr.gain)
  
 
  
  loss.matrix <- get(outname.enrichr.loss)
  
  
  
  if(nrow(gain.matrix)>0){
    gain.matrix$Signature <- signature
    colnames(pathways.summary.gain.binding) <- colnames(gain.matrix)
    
    pathways.summary.gain.binding <- rbind(pathways.summary.gain.binding,gain.matrix)
    
  }
  
  if(nrow(loss.matrix)>0){
    loss.matrix$Signature <- signature
    colnames(pathways.summary.loss.binding) <- colnames(loss.matrix)
    
    pathways.summary.loss.binding <- rbind(pathways.summary.loss.binding,loss.matrix)
    
  }
  
  
  
}



combined.pathways <- unique(c(pathways.summary.gain.binding$New.Term,pathways.summary.loss.binding$New.Term))

combined.dot.matrix.plot <- data.frame(matrix(nrow=length(combined.pathways),ncol=47))

row.names(combined.dot.matrix.plot) <- combined.pathways

colnames(combined.dot.matrix.plot) <- unartifact.signatures


combined.dot.matrix.plot[,] <- 0

for(i in 1:nrow(combined.dot.matrix.plot)){
  
  if(length(which(pathways.summary.gain.binding$New.Term==
                  row.names(combined.dot.matrix.plot)[i]))>0){
    
    signature.list <- pathways.summary.gain.binding$Signature[which(pathways.summary.gain.binding$New.Term==
                                                                      row.names(combined.dot.matrix.plot)[i])]
    
    combined.dot.matrix.plot[i,signature.list] <- combined.dot.matrix.plot[i,signature.list] + 1
    
  }
  
  if(length(which(pathways.summary.loss.binding$New.Term==
                  row.names(combined.dot.matrix.plot)[i]))>0){
  
  signature.list <- pathways.summary.loss.binding$Signature[which(pathways.summary.loss.binding$New.Term==
                                                                    row.names(combined.dot.matrix.plot)[i])]
  
  combined.dot.matrix.plot[i,signature.list] <- combined.dot.matrix.plot[i,signature.list] + 2
  }
  
  
}

summary.matrix <- data.frame(combined.pathways)

for(i in 1:nrow(summary.matrix)){
  
  summary.matrix$freq[i] <- sum(combined.dot.matrix.plot[i,]!=0)
}

specific.pathway.order <- summary.matrix$combined.pathways[order(summary.matrix$freq,decreasing = T)]


plot.dot.matrix <- melt(as.matrix(combined.dot.matrix.plot))
colnames(plot.dot.matrix) <- c("Pathway","Signature","value")

mycol <- c("grey","tomato1", "dodgerblue", "darkorchid3")




sum.of.pathways.both <- data.frame(unartifact.signatures)
sum.of.pathways.gain <- data.frame(unartifact.signatures)
sum.of.pathways.loss <- data.frame(unartifact.signatures)



for(i in 1:length(unartifact.signatures)){
  
  sum.of.pathways.both$freq[i] <- sum(combined.dot.matrix.plot[,i]==3)
  sum.of.pathways.loss$freq[i] <- sum(combined.dot.matrix.plot[,i]==2)
  sum.of.pathways.gain$freq[i] <- sum(combined.dot.matrix.plot[,i]==1)
  
}

for(i in 1:nrow(sum.of.pathways.both)){
  
  signature <- sum.of.pathways.both$Signature[i]
  
  selected.matrix <- PCAWG_sigProfiler_SBS_signatures_in_samples[which(PCAWG_sigProfiler_SBS_signatures_in_samples[,signature]!=0),]
  
  sum.of.pathways.both$type.of.cancer[i] <- length(unique(selected.matrix$Cancer.Types))
}

sum.of.pathways.both$Category <- "Both"
sum.of.pathways.loss$Category <- "Loss"
sum.of.pathways.gain$Category <- "Gain"

colnames(sum.of.pathways.both) <- c("Signature","Freq","Category")
colnames(sum.of.pathways.loss) <- c("Signature","Freq","Category")
colnames(sum.of.pathways.gain) <- c("Signature","Freq","Category")

sum.of.pathways <- rbind(sum.of.pathways.both,sum.of.pathways.loss,sum.of.pathways.gain)

for(i in 1:nrow(sum.of.pathways)){
  
  sum.of.pathways$Sum[i] <- sum(sum.of.pathways$Freq[which(sum.of.pathways$Signature==sum.of.pathways$Signature[i])])
  
  
}
specific.signature.order <- unique(sum.of.pathways$Signature[order(sum.of.pathways$Sum,decreasing=T)])





```


